A Test of the Frequency Independence Assumption of the Power System Parameters used in Geomagnetically Induced Current Calculations

### 2019-02-23 paper.tex
Updates from Overleaf

\begin{keypoints}
\item The power system parameters derived from GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependence significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used in estimating geomagnetically induced currents (GICs) in a power system given a direct measurement or indirect estimate (via a transfer function with input of the geomagnetic field) of the horizontal geoelectric field components $E_x$ and $E_y$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  We provide the first test of this assumption using GIC measurements made in a 187 kV transformer connected to a $\sim$100 km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$10\todo[color=blue!40]{Shinichi: Is 10 km correct?} kilometers away.  We compute frequency dependent network parameters $A(\omega)$ and $B(\omega)$ in Model 2, $GIC(\omega)$ $=$ $A(\omega)E_x(\omega) + B(\omega)E_y(\omega)$, and find that they are frequency dependent and that this model provides a significantly better representation of measured GIC.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems due to electric fields induced near Earth's surface. These fields are a result of time variations in ionospheric currents on timescales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs given direct measurements of the electric field (or estimates of the electric field based on the geomagnetic field) is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \todo{Pierre and Trevor - any suggestions?} {\color{red} \textbf{[REFS]}}.

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression, when contemporaneous measurements of GIC and $\mathbf{E}$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Viljanen2012,Overbye2012}.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals: 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals cover those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1-s raw data and 1 Hz low-pass filtered raw measurements. The results presented here are for the 1 Hz low-pass filtered GIC values

One-second-cadence $\mathbf{E}$ and $\mathbf{B}$ measurements a from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal. Data from 2001-2011 have been used by ~\cite{Fujii2015} to estimate the magnetotelluric impedance at the site.

Data for the longest continuous time span of GIC measurements, 2006-04-02 through 2006-04-10 are shown Figure~\ref{sample}.

@@

The only modification made to any of the magnetic field measurements was the replacement of 36 data points in all components of the magnetic field on 2006-08-06 with linearly interpolated values. The values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

The GIC data contains non-physical spikes that are followed by a relaxation time of 40-80 s. These spikes were removed by identifying times $t_s$ when the change in 1-se $|GIC|$ measurements was greater than $0.04$ Amps and then data in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ were replaced using a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For all of the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

To determine the coefficients $a_o$ and $b_o$ in the frequency independent model, Model ~1,

@@

\noindent
the matrix equation $\mathbf{GIC} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is created, where $\mathbf{GIC}$ is a 864000x1 matrix of 1 day of observations, $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ observations. The values of $\mathbf{p}$ are then obtained using a standard linear regression solver. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][ provided the mathematically equivalent closed-form equations.]{Pulkkinen2007}. The parameters $a_o$ and $b_o$ were computed for each of the 35 unique 1-day window segments of available data.

In Model 2 the system parameters are frequency dependent

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)\,.
\label{model2}
\end{equation}
\end{linenomath*}

The third model considered uses an electric field $\mathbf{E}'$ in Equation~\ref{model2}, where $\mathbf{E}'$ was computed using the measured magnetic field and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) = \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$:

@@

\noindent
where, explicitly, the electric field components are $E'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $E'_y = Z_{yx}B_x + Z_{yy}B_y$. Substitution gives

\begin{equation*}
G_{E'}(\omega) = \big[a(\omega)Z_{xx}(\omega) + b(\omega)Z_{yx}(\omega) \big] B_x(\omega) + \big[ a(\omega)Z_{xy}(\omega) + b(\omega)Z_{yy}(\omega) \big] B_y(\omega)\,.
\end{equation*}

That is, instead of using the measured electric field directly as in Model 2, a transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field. In this work, $\boldsymbol{Z}$ was calculated using the method described later in this section; $a(\omega)$ and $b(\omega)$ are taken from the solution of Equation~\ref{model2}, using the method also described later in this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of GIC and $\mathbf{B}$ and the model is expected to produced predictions of GICs that are similar to that obtained from Model 3.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e = 2\pi f_e$, by performing a least squares regression using Equation~\ref{model2} and the set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the fourier transform of 24 hours of 1-second data. The values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$; if $f_e$ value is not a multiple of $1/86400$, the nearest multiple is chosen. The regression for $f_e$ uses all $f$ values between the next higher and lower value of $f_e$.

To convert the frequency domain transfer functions $a(\omega)$ and $b(\omega)$ estimated at the frequencies $\omega_e$ to impulse responses $a(\tau)$ and $b(\tau)$, they are first linearly interpolated onto a 1 Hz frequency grid and then inverse fourier transformed. The transfer function magnitudes $|a(\omega)|$ and $|b(\omega)|$ were calculated by linear interpolation of the transfer function magnitudes (instead of linearly interpolating the real and imaginary components and then computing the magnitude). The same approach is used for solving for $z_x$ and $z_y$ in Equation~\ref{model4}. For the transfer function $\boldsymbol{Z}$ in $\mathbf{E}(\omega) = \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$, the $xx$ and $yx$ components of $\boldsymbol{Z}$ are solved for using measured values of $E_x$, $B_x$, and $B_y$ in $E_x = Z_{xx}B_x + Z_{yx}B_{y}$; the $yx$ and $yy$ components are solved for using measured values of $E_y$, $B_x$, and $B_y$ in $E_y = Z_{yx}B_x + Z_{yy}B_y$.

@@

There are many statistical methods that have been developed for estimating transfer functions in the magnetotelluric literature that use additional steps and processing \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used that follow the basic method in \cite{Simpson2005}. These works typically involve the estimation of a robust and bias-corrected transfer function. In the results presented, we did not use a robust statistical method or a bias correction algorithm as our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating the model parameters. As discussed in~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model (when the sum-of-squares error is used as a prediction metric).

However, several alternative methods were tested individually by modifying the basic method described above to determine if there was a change in the computed transfer functions and, most important for this work, the out-of-sample prediction performance of each model. When using these alternative methods, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are generally within the error bars of the used method when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s). The experiments in this regard are discussed in the Appendix.

@@

The performance of each model in predicting GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is defined as PE = 1 - variance(Prediction Error)/variance(Predicted); it is a measure of the skill of the model relative to the variance in the data that are predicted, and it represents the fraction of the variance in the predicted data that predicted by the model.

The average PE and MSE ratio was computed by averaging the 35 out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the out-of-sample 35 signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the 35 prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

@@

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model 2, is significantly better at predicting GIC than the frequency independent model, Model 1. Model 3, which uses magnetic field measurements as in input provides better estimates of GIC than Model 2. Model 4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides the best predictions of GIC.

Figure~\ref{SN} shows the average signal-to-noise ratios as a function of period for the four models. The error bars represent $\pm$ one standard error of the 35 SN ratios at each period. Consistent with the results in Table~1, The SN ratio for Model 1 is lower than that for Models 2-4 at all frequencies.  Above $1400$~s (24~min), the SN ratio ordering is consistent with the ordering of the PE and MSE ratios shown in Table~1, with Model~1 having the lowest and Model~4 having the highest SN ratio. Below $500$~s (8~min), Model 2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high frequency fluctuations in the prediction error for Model 2 are smaller than that for Models 3 and 4.

@@

The wide distribution and large error bars are suggestive that Model~\ref{model1} is a poor representation of the measurements. However, the model does have a good prediction performance - the PE is 0.35 and the data/model correlation coefficient of 0.64. In a similar way that $|\mathbf{B}(t)|$ or $d|\mathbf{B}(t)|/dt$ are often good predictors of $GIC(t)$  \todo{Pierre: Any suggestions?} {\color{red} \textbf{[REFS]}} despite the fact that physically $\mathbf{E}$ drives GIC and $\mathbf{E}$ is related to $\mathbf{B}$ through a frequency dependent transfer function, Model 1 is a fair predictor of GICs, but is a poor approximation relative to Models 2 and 3. For the data considered in this work, when a model $G(t) = $ $c_oB_x(t) + d_oB_y(t)$ is used, were $c_o$ and $d_o$ are constants, the in-sample average prediction efficiency is comparable to that of Model 1, but the out-of-sample prediction efficiency is negative.

The average parameters for Models 2 and 4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average.

Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model 2 and $\mathbf{B}$ for Model 4 are similar and are different from zero at more than one time value (the impulse response for Model 3 is similar to that of Model 2 and is not shown). By definition, the impulse response for Model 1 is zero except at $t=0$. Note that the sign of the impulse response at $t=0$ is opposite for Model 1 and Model 2 for the $E_y$ coefficient - at $t=0$, $b_o < 0$ while $b(0)>0$.

Figure~\ref{Z} shows that the frequency domain transfer functions for Model 2 vary by an order of magnitude on timescales of 60 seconds to 12 hours (that for Model 1 is constant by definition). Note that the ratio of $|\overline{b}_o/||\overline{a}_o|$ is approximately 1 for Model~1 but varies from 1.6 to 6.3 for Model~2. In Model~2, GIC is most dependent on $E_y$, however note that $E_y$ has less variability than $E_x$ - the variance in $E_x$ is $\sim$5x larger than $E_y$ over the entire dataset (this is also visible in Figure~\ref{sample}).

Figure~\ref{Phi} shows the transfer function phases. By definition, that for Model 1 is zero. The phases for $z_x$ and $b$ are offset by 120 degrees \todo{Why?} while the phases for $a$ and $b$ are significantly different \todo{Why?}.

@@

Several methods have been used in the literature for estimating GIC. When electric field measurements are available, a model of the form of Model~1 is used \todo{TODO} {\color{red} \textbf{[REFS]}}. We have shown using a data set of contemporaneous GIC and $\mathbf{E}$ measurements that a significantly better estimation of GIC can be obtained by using a model of the form of Model~2.

When only a MT impedance tensor that connects $\mathbf{E}$ to $\mathbf{B}$ is available, a model of the form of Model~1, with the electric field estimated using $\mathbf{E}=\mathbf{Z}\mathbf{B}$ \todo{TODO} {\color{red} \textbf{[REFS]}}. For the dataset considered, significantly better estimates can be made if frequency dependent system parameters are available as in Model~2.

When only magnetic field data is available, a model of the form of Model~4 is used. In this case, the transfer function contains components of the transfer function that connects $\mathbf{E}$ to $\mathbf{B}$ and GIC to $\mathbf{E}$.

\todo{TODO} {\color{red}\textbf{Discuss the origin of the frequency dependence of the system parameters.}}

>>>>

\begin{keypoints}
\item Power system parameters derived from a set of GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependence significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used in estimating geomagnetically induced currents (GICs) in a power system given a nearby direct measurement or indirect estimate (via a transfer function with input of the geomagnetic field) of the horizontal geoelectric field components $E_x$ and $E_y$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  We provide the first test of this assumption using GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$10~km \todo[color=blue!40]{Shinichi: Is 10 km correct?} away.  We compute frequency dependent network parameters $a(\omega)$ and $b(\omega)$ in Model 2, $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$, and find that they are frequency dependent and that this model provides a significantly better representation of measured GIC.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems due to electric fields induced near Earth's surface. These fields are a result of time variations in ionospheric currents on timescales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs given direct measurements of the electric field (or estimates of the electric field based on the geomagnetic field) is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \citep{Molinski2002,Thomson2010,NERC2012,Gaunt2014}. \todo{Pierre and Trevor - any other reference suggestions?}

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression, when contemporaneous measurements of GIC and $\mathbf{E}$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Horton2012,Viljanen2012,Overbye2012,Marshall2013,Liu2014,Zheng2014,Watari2015,Bonner2017}.

\cite{Oyedokun2013a} developed a method to account for the finite inductive response time of transformers by taking the GIC predicted based on the quasi-DC assumption and then applying a piecewise-linear L/R filter derived from bench-scale transformer measurements, with a time constant dependent on load, to obtain an alternative estimate of GIC.

\todo[inline]{What can I conclude from \cite{Oyedokun2013a}? Mostly the thesis documents visual differences in time series traces with and without accounting for transformers.}

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST): 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals cover those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass filtered raw measurements. The results presented here are for the 1~Hz low-pass filtered GIC values

One-second-cadence $\mathbf{E}$ and $\mathbf{B}$ measurements a from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal.

Data for the time span of 2006-08-04T15:00 through 2006-08-08T15:00 are shown Figure~\ref{sample}. The GIC data were provided as 1-day files that start at midnight JST and the times shown in Figure~\ref{sample} are Universal Time.

@@

The only modification made to the magnetic field measurements was the replacement of 36 data points in all components of the magnetic field on 2006-08-06 with linearly interpolated values. The values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

The GIC data contains non-physical spikes that are followed by a relaxation time of 40-80 s. These spikes were removed by identifying times $t_s$ when the change in 1~s $|GIC|$ measurements was greater than $0.05$ Amps and then data in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ were replaced using a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For all of the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

To determine the coefficients $a_o$ and $b_o$ in the frequency independent model, Model~1,

@@

\noindent
the matrix equation $\mathbf{GIC} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is created, where $\mathbf{GIC}$ is a 864000x1 matrix of 1 day of observations, $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ observations. The values of $\mathbf{p}$ are then obtained using a standard linear regression solver. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations.]{Pulkkinen2007}. The averages $\overline{a}_o$ and $\overline{b}_o$ are the average of the 35 $a_o$ and $b_o$ values computed using the 35 unique 1-day window segments of available data.

In Model 2, the system parameters $a$ and $b$ are frequency dependent according to

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are obtained using contemporaneous measurements of GIC and $\mathbf{E}$ using the method described later in this section. Note that fourier transformed variables are indicated by explicitly listing their dependence on $\omega$.

In Model~3, the parameters $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)/\mu_o$:

@@

\noindent
where, explicitly, the electric field components are $\mu_oE'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $\mu_oE'_y = Z_{yx}B_x + Z_{yy}B_y$. Substitution gives

\begin{linenomath*}
\begin{equation*}
G_{E'}(\omega) = \big[a(\omega)Z_{xx}(\omega) + b(\omega)Z_{yx}(\omega) \big] B_x(\omega)/\mu_o + \big[ a(\omega)Z_{xy}(\omega) + b(\omega)Z_{yy}(\omega) \big] B_y(\omega)/\mu_o\,.
\end{equation*}
\end{linenomath*}

That is, instead of using the measured electric field directly as in Model 2, a transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field. In this work, $\boldsymbol{Z}$ was calculated using the method described later in this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of GIC and $\mathbf{B}$ and the model is expected to produce predictions of GICs that are similar to that obtained from Model 3.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a least squares regression using Equation~\ref{model2} and the set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the fourier transform of 24 hours of 1-s cadence data. The values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$; if $f_e$ value is not a multiple of $1/86400$, the nearest multiple is chosen. The regression for $f_e$ uses all $f$ values between the next higher and lower value of $f_e$.

To convert the frequency domain transfer functions $a(\omega)$ and $b(\omega)$ estimated at the frequencies $\omega_e$ to impulse responses $h_a(\tau)$ and $h_b(\tau)$, their real and imaginary components are linearly interpolated onto a 1 Hz frequency grid and then inverse fourier transformed.

The same approach is used for solving for $z_x$ and $z_y$ in Equation~\ref{model4}. For the transfer function $\boldsymbol{Z}$ in $\mathbf{E}(\omega) = \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ used in Model~3, the $xx$ and $yx$ components of $\boldsymbol{Z}$ are solved for using $E_x(\omega) = Z_{xx}(\omega)B_x(\omega) + Z_{yx}(\omega)B_{y}(\omega)$; the $yx$ and $yy$ components are solved for using $E_y(\omega) = Z_{yx}(\omega)B_x(\omega) + Z_{yy}(\omega)B_y(\omega)$.

@@

There are many statistical methods and variations thereof that have been developed for estimating transfer functions in the magnetotelluric (MT) literature that use additional steps and processing \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used that follow the basic method in \cite{Sims1971,Simpson2005}. These objective of these works is typically the estimation of a robust and low bias transfer function. In the results presented, we did not use a robust statistical method or attempt to correct for bias in estimates as our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating the model parameters. As discussed and evaluated in~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model (when the sum-of-squares error is used as a prediction metric).

However, several additional processing methods and steps used in the MT literature were tested individually by modifying the basic least-squares method used to determine if there was a change in the computed transfer functions and, most relevant for this work, the out-of-sample prediction performance of each model. When using these additional steps, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are generally within the error bars of the used method when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s). The experiments in this regard are discussed in the Appendix.

@@

The performance of each model in predicting GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio. 
The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is defined as PE = 1 - variance(Prediction Error)/variance(Predicted); it is a measure of the skill of the model relative to the variance in the data that are predicted, and it represents the fraction of the variance in the predicted data that predicted by the model.

The average PE and MSE ratio was computed by averaging the 35 out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the 35  out-of-sample signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the 35 prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

@@

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model 2, is significantly better at predicting GIC than the frequency independent model, Model 1. Model 3, which uses horizontal magnetic field measurements as its input provides better estimates of GIC than Model 2. Model 4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides the best predictions of GIC.

Figure~\ref{SN} shows the average signal-to-noise ratios as a function of period for the four models. The error bars represent $\pm$ one standard error of the 35 SN ratios at each period. At each evaluation frequency the signal-to-noise ratio is is the ratio of the average of the 35 segment PSDs of GIC to the average of the 35 segment PSDs of the prediction error. The PSD for each of the segments at a given evaluation frequency was obtained by averaging the raw PSD using values between the next higher and lower evaluation frequency.

Consistent with the results in Table~1, The SN ratio for Model 1 is lower than that for Models 2-4 at all frequencies.  Above $10,000$~s (2.8~hrs), the SN ratio ordering is the same as the ordering of the PE and MSE ratios shown in Table~1, with Model~1 having the lowest and Model~4 having the highest SN ratio. Below $500$~s (8~min), Model 2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high frequency fluctuations in the prediction error for Model 2 are smaller than that for Models 3 and 4.

@@

The wide distribution of parameter values and large error bars are suggestive that Model~\ref{model1} is a poor representation of the measurements. However, the model does have a good prediction performance - the PE is 0.35 and the data/model correlation coefficient of 0.64. In a similar way that $|\mathbf{B}|(t)$ or $d|\mathbf{B}|(t)/dt$ are often good predictors of $GIC(t)$ \todo{Pierre: Any suggestions?} {\color{red} \textbf{[REFS]}} despite the fact that physically $\mathbf{E}$ drives GIC and $\mathbf{E}$ is related to $\mathbf{B}$ through a frequency dependent transfer function, Model 1 is a fair predictor of GICs, but is a poor approximation relative to Models 2 and 3.

%However, for the data considered in this work, when a model $G(t) = $ $c_oB_x(t) + d_oB_y(t)$ is used, were $c_o$ and $d_o$ are constants, the in-sample average prediction efficiency is comparable to that of Model 1, but the out-of-sample prediction efficiency is negative.

The average parameters for Models 2 and 4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average. The transfer function magnitudes were calculated by linear interpolation of the transfer function magnitudes at the evaluation frequencies (instead of linearly interpolating the real and imaginary components and then computing the magnitude).

Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model 2 and $\mathbf{B}$ for Model 4 are similar and are different from zero at more than one time value (the impulse response for Model 3 is similar to that of Model 2 and is not shown). By definition, the impulse response for Model 1 is zero except at $t=0$. Note that the sign of the impulse response at $t=0$ is opposite for Model 1 and Model 2 for the $E_y$ coefficient - at $t=0$, $b_o < 0$ while $h_b(t=0)>0$. An explanation for the acausality of the impulse response is given in \cite{Kelbert2017}.

Figure~\ref{Z} shows that the frequency domain transfer functions for Model 2 vary by an order of magnitude on timescales of 60 seconds to 12 hours (that for Model 1 is constant by definition). In Model~2, GIC more sensitive to $E_y$ due to $b(\omega) > a(\omega)$, but note that $E_y$ has less variability than $E_x$ - the variance in $E_x$ is $\sim$5x larger than $E_y$ over the entire dataset (this is also visible in Figure~\ref{sample}). The ratio of $|\overline{b}_o/|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h. At $T=12$~h, the highest period considered, the phases of $a(\omega)$ and $b(\omega)$ shown in Figure~\ref{Phi} differ by 180 degrees and the signs are consistent with $\overline{a}_o$ and $\overline{b}_o$.

Figure~\ref{Phi} shows the transfer function phases. By definition, that for Model 1 is zero. The phases for $z_x(\omega)$ and $b(\omega)$ are offset by 120 degrees while the phases for $a(\omega)$ and $b(\omega)$ are significantly different. \todo{Why?} At a period of $60$~s, the parameters in Equation~\ref{model2} are most consistent with the assumptions of frequency independent system parameters of Model~1. The phases of $a(\omega)$ and $b(\omega)$ are near zero and the $b(\omega)$ curve is the flattest.

%\cite{Pulkkinen2010} noted that one could estimate the ratio of $a_o/b_o$ using GIC and local geomagnetic field measurements even if $\mathbf{Z}(\omega)$ was unknown if one assumed the ground conductivity was 1-D so that $Z\equiv Z_{xy}=-Z_{yx}$ and $Z_{xx}=Z_{yy}=0$ so that Equation~\ref{model2} can be written

%\begin{linenomath*}
%\begin{equation}
%G(\omega) = a_oZ(\omega)B_y(\omega)/\mu_o - b_oZ(\omega)B_x(\omega)/\mu_o\,.
%\label{pulkkinen2010}
%\end{equation}
%\end{linenomath*}

%\noindent
%If the above equations are solved for with unknowns $Z_a\equiv a_oZ$ and $Z_b\equiv b_oZ$,  then $a_o/b_o = Z_a(\omega)/Z_b(\omega)$. \cite{Pulkkinen2010} assumed that this ratio was frequency independent and provided a closed-form equation for $a_o/b_o$ by multiplying Equation~\ref{pulkkinen2010} by the complex conjugates of $B_x(\omega)$ and $B_y(\omega)$ so as to form two equations from which $Z$ can be eliminated. With the additional assumption that the cross- and auto-spectra of all variables have the same functional form, they showed that the ratio can be written in a way that requires only evaluation of correlations among $G(t)$, $B_x(t)$, and $B_y(t)$.

@@

The primary implications of the results presented are (1) Significantly improved predictions of GICs can be made if frequency dependent system parameters are used and (2) A direct comparison of $a_o$ and $b_o$ estimated using Equation~\ref{model1} and computed using power system configuration information is problematic because the data are not consistent with the assumptions of Equation~\ref{model1}. Although the use of $a_o$ and $b_o$ values estimated from the data give a reasonable out-of-sample prediction model, it is not known how much these data estimated values will differ from those directly calculated.

Also note that $\sim 10$ days of data are required to give values of $a_o$ and $b_o$ that provide the out-of-sample prediction quality reported here for Model~1 (a PE of 0.35 and CC of 0.64). If only a few days of data are used for their estimation, the out-of-sample prediction quality is very poor due to the high degree of uncertainty of in their estimates. Predictions of GIC using Model~1 in the literature using only a few days of data are expected to be significantly worse if out-of-sample data are used.

Several methods have been used in the literature for estimating or predicting GIC depending on the data or models that are available.

When electric field measurements are available, a \todo{TODO} {\color{red} \textbf{[REFS]}} model of the form of Model~1, with frequency independent system parameters derived from power system configuration, has been used. These works generally note a high degree of visual consistency between the measured values and predictions using this model (e.g., the top panel of Figure~\ref{predictions}). However, we have shown using a data set of contemporaneous GIC and $\mathbf{E}$ measurements that a significantly better estimation of GIC can be obtained by using a frequency dependent model of the form of Model~2. In addition, we have found that the estimates of the system parameters $a_o$ and $b_o$ are highly sensitive to the interval of data used, and have a high degree of uncertainty.

When only a MT impedance tensor that connects $\mathbf{E}$ to $\mathbf{B}$ is available, a model of the form of Model~1, with the electric field estimated using $\mathbf{E}=\mathbf{Z}\mathbf{B}$ has been used~\citep{Weigel2017,Bonner2017,Kelbert2017}. For the dataset considered, significantly better GIC estimates can be made if frequency dependent system parameters are available as in Model~3.
